---
title: "Performance comparison of crr and crprep + coxph"
author: "Alessandro Gasparini"
date: "`r Sys.time()`"
output: html_document
---
```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, eval = FALSE)
```
# gen_cr_data( )

```{r gen_cr_data}
gen_cr_data <- function(n = 1000){
  require(dplyr)
  data <- data_frame(time0 = rexp(n, 0.1),
                     time1 = rexp(n, 0.5),
                     time2 = rexp(n, 1)) %>%
    mutate(time = round(1 + 100 * pmin(time0, time1, time2)),
           status = ifelse(time1 == pmin(time0, time1, time2),
                           1,
                           ifelse(time2 == pmin(time0, time1, time2),
                                  2,
                                  0)),
           trt = sample(0:1, n, replace = TRUE),
           age = abs(rnorm(n, 50, 10))) %>%
    select(-time0, -time1, -time2)
  return(tbl_df(data))
}
```

This function generates a dataset with 2 competing events, a factor with two levels `trt` and a continuous variable `age`. Sample size is customizable by the user changing the argument `n` (default value is 1.000).

Use `set.seed()` for reproducibility.

# crr_vs_crprep1( )

```{r crr_vs_crprep1}
crr_vs_crprep1 <- function(ssize, B = 1000){
  if (length(ssize) != 1) stop("ssize cannot be a vector.")
  require(pacman)
  p_load("cmprsk", "mstate", "microbenchmark")
    d <- gen_cr_data(round(ssize))
    mb <- microbenchmark(
      "CRR" = with(d, crr(ftime = time, fstatus = status, cov1 = cbind(age, trt), tf = function(t) t, failcode = 1, cencode = 0)),
      "CRPREP + COXPH" = with(with(d, crprep(Tstop = time, status = status, data = d, trans = 1:2, cens = 0, keep = cbind(age, trt))),
                              coxph(Surv(Tstart, Tstop, status == 1) ~ age + trt, weights = weight.cens, subset = (failcode == 1))),
      times = B
      )
    return(mb)
}
```

This function compares the time necessary to estimate a Fine & Gray model using the standard estimator versus the time necessary to compute weights and consequently a weighted Cox semiparametric model, as suggested in [1], using the `microbenchmark()` function from the homonymous package [2].

This function works with a single sample size at a time, and it is possible to set the number of replications with the `B` argument (defaul of 1.000 replications).

To obtain time estimates for different sample sizes, use `lapply` (`lapply(c(10, 100, 1000), crr_vs_crprep1)`) or the least efficient `for` cycle.

# crr_vs_crprep2( )

First, define two separate functions for the two procedures:

```{r crr_vs_crprep2_1}
# Regular
f1 <- function(d, M){
        with(d, lapply(1:M, function(x) crr(ftime = time, fstatus = status, cov1 = cbind(age, trt), tf = function(t) t, failcode = 1, cencode = 0)))
}
# Weighted
f2 <- function(d, M){
          d_wt <- with(d, crprep(Tstop = time, status = status, data = d, trans = 1:2, cens = 0, keep = cbind(age, trt)))
          with(d_wt, lapply(1:M, function(x) coxph(Surv(Tstart, Tstop, status == 1) ~ age + trt, weights = weight.cens, subset = (failcode == 1))))
      }
```

Then, compare the two:

```{r crr_vs_crprep2_2}
crr_vs_crprep2 <- function(M, ssize = 100, B = 1000){
  if (length(ssize) != 1) stop("ssize cannot be a vector.")
  require(pacman)
  p_load("cmprsk", "mstate", "microbenchmark", "parallel")
    d <- gen_cr_data(round(ssize))
    mb <- microbenchmark(
      "CRR" = f1(d, M),
      "CRPREP + COXPH" = f2(d, M),
      times = B
      )
    return(mb)
}
```

This function compares the time necessary to estimate `M` Fine & Gray models using the standard `crr()` function from the `cmprsk` package versus the time necessary to estimate the same number of models restructuring the data first (`crprep` function) and then using the weighted Cox regression model.

This function works with a single sample size at a time, and it is possible to set the number of models to estimate with the `M` argument(default to 100 models) and the number of replications with the `B` argument (defaul to 1.000 replications).

To obtain time estimates for different sample sizes act analogously to `crr_vs_crprep1()`.

# Usage

## Simulation 1

```{r usage1}
require(pacman)
p_load("ggplot2", "parallel", "ggthemes")
ssizes <- c(50, 100, 500, 1000, 5000, 10000)
uno <- lapply(ssizes, crr_vs_crprep1, B = 1000)
jnd <- bind_rows(mclapply(1:length(ssizes), 
                          function(x){ 
                            return(cbind(uno[[x]], ssize = ssizes[x]))
                            }))

ggplot(data = jnd, aes(x = expr, y = time)) + geom_boxplot() + facet_wrap(~ ssize, nrow = round(length(ssizes) / 3)) + labs(title = "", x = "", y = "Time: log10(ms)") + scale_y_log10() + theme_few()

for(i in ssizes){
  print(i)
  print(with(filter(jnd, ssize == i), wilcox.test(time ~ expr)))
}
```

## Simulation 2

```{r usage2}
require(pacman)
p_load("ggplot2", "parallel", "ggthemes")
ssizes <- c(100, 1000, 10000)
emmes <- c(10, 50, 100)
bi <- 1000

#1st Sample Size
dos1 <- lapply(emmes, crr_vs_crprep2, ssize = ssizes[1], B = bi); dos1
jnd <- bind_rows(mclapply(1:length(emmes),
                            function(x){ 
                              return(cbind(dos1[[x]], emme = emmes[x]))
                              }))
jnd1 <- mutate(jnd, ssize = ssizes[1])

#2nd Sample Size
dos2 <- lapply(emmes, crr_vs_crprep2, ssize = ssizes[2], B = bi); dos2
jnd <- bind_rows(mclapply(1:length(emmes),
                            function(x){ 
                              return(cbind(dos2[[x]], emme = emmes[x]))
                              }))
jnd2 <- mutate(jnd, ssize = ssizes[2])

#3rd Sample Size
dos3 <- lapply(emmes, crr_vs_crprep2, ssize = ssizes[3], B = bi); dos3
jnd <- bind_rows(mclapply(1:length(emmes),
                            function(x){ 
                              return(cbind(dos3[[x]], emme = emmes[x]))
                              }))
jnd3 <- mutate(jnd, ssize = ssizes[3])

#Combine
jnd_t <- bind_rows(jnd1, jnd2, jnd3) %>%
  mutate(fct = paste("n =", ssize, "| m =", emme))

# jnd_t$fct = factor(jnd_t$fct, levels = c("n = 100 | m = 10", "n = 100 | m = 50", "n = 100 | m = 100", "n = 1000 | m = 10", "n = 1000 | m = 50", "n = 1000 | m = 100", "n = 10000 | m = 10", "n = 10000 | m = 50", "n = 10000 | m = 100"))

for(i in levels(factor(jnd_t$fct))){
  print(i)
  print(with(filter(jnd_t, fct == i), wilcox.test(time ~ expr)))
}

ggplot(data = jnd_t, aes(x = expr, y = time)) + geom_boxplot() + facet_wrap(~ fct, nrow = length(ssizes)) + labs(title = "", x = "", y = "Time: log10(ms)") + scale_y_log10() + theme_few()
```

# References

[1] R. B. Geskus, _Cause-specific cumulative incidence estimation and the Fine and Gray model under both left truncation and right censoring_, Biometrics, 67, 39–49 (2011)

[2] `microbenchmark` package on [CRAN](http://cran.r-project.org/web/packages/microbenchmark/microbenchmark.pdf)

[3] `parallel` package on [CRAN](https://stat.ethz.ch/R-manual/R-devel/library/parallel/doc/parallel.pdf)